---
title: 'Inferential Statistics: Exploration of the Stroop Effect'
author: "Tyler C Julian"
date: "February 19, 2018"
header-includes:
  - \usepackage{mathtools}
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Introduction

The Stroop Effect, named after the original researcher John Ridley Stroop, is a psychological phenomenon that demonstrates the effect of interference on reaction time. The basis of the experiment consists of participants reading aloud  the color of each word from a list of colored words as fast and correctly as possible, with their final time being recorded. There are two trials, each with a unique condition: the first where the words are congruent and the second where the words are incongruent. _Congruent words_ are words where the content of the word matches the color of the word: for example, <span style="color:red">RED</span>, <span style="color:green">GREEN</span>, <span style="color:blue">BLUE</span>. _Incongruent words_ are words where the content of the word does not match the color: for example <span style="color:green">RED</span>, <span style="color:orange">PURPLE</span>, <span style="color:blue">PINK</span>. The amount of time it takes to complete each trial is measured and recorded for each participant. 

In this project, the Stroop Effect will be explored again, but with a focus on the statistical context. The experiment will be run using a sample consisting of a provided dataset and my own recorded times. The independent and dependent variables of this study will be identified and a hypothesis testing will also be conducted. Ultimately, the sample dataset will undergo a statistical test, and a conclusion will be reached based on the results. 

### Independent and Dependent variables

The __independent variable__ for this experiment is the conditions of the words. They change from congruent to incongruent between the two trials.

The __dependent variable__ is the amount of time it takes a participant to complete each trial, measured in seconds. 

### Hypothesis and Statistical Test

The __null hypothesis__ for this experiment is that there will not be a statistical difference in the average time between the congruent and incongruent conditions. 

The __alternative hypothesis__ is that there will be a statistical difference in the average time between the congruent and incongruent conditions.

These hypotheses can be pictured mathematically below, with $H_0$ being the null hypothesis, $H_a$ being the alternative hypothesis, $\mu_c$, being the average congruent time, and $\mu_i$ being the average incongruent time:

$$
H_0 : \mu_c = \mu_i \\
H_a : \mu_c ≠ \mu_i
$$

The statistical test being used for this experiment is a __two-tailed paired sample t-test__. The __confidence level__ will be set to  _0.05_. This t-test is a dependent t-test with paired samples and is chosen because the population parameters are unknown. It is two-tailed because we are unsure whether the times will rise or fall between word conditions.

### Loading the Dataset into R and Running the Experiment

The starting dataset consists of 24 paired observations to start with. This sample of records was retrieved from [this site](https://drive.google.com/file/d/0B9Yf01UaIbUgQXpYb2NhZ29yX1U/view), provided by Udacity.

```{r, Read Data}
congruent <- c(12.079, 16.791, 9.564, 8.63, 14.669, 12.238, 14.692, 8.987,
               9.401, 14.48, 22.328, 15.298, 15.073, 16.929, 18.2, 12.13,
               18.495, 10.639, 11.344, 12.369, 12.944, 14.233, 19.71, 16.004)

incongruent <- c(19.278, 18.741, 21.214, 15.687, 22.803, 20.878, 24.572, 17.394,
                 20.762, 26.282, 24.524, 18.644, 17.51, 20.33, 35.255, 22.158,
                 25.139, 20.429, 17.425, 34.288, 23.894, 17.96, 22.058, 21.157)

stroop <- data.frame(congruent, incongruent)
```

With the bulk of the sample loaded into R, I now need to run the experiment myself and add my times to the dataset.

If you are interested, I encourage that you run the experiment yourself as well! [It is hosted by the University of Washington and is free to use](https://faculty.washington.edu/chudler/java/ready.html). It definitely helped me better understand the design of the experiment (plus, I found it quite fun).

After running the test, I scored __10.278__ secs on my first (congruent) run and __18.169__ secs on my second (incongruent) run. I can now add my own results to the dataset: 

```{r Add Personal Times}
stroop <- rbind(stroop, c(10.278, 18.169))
```

With my own data now in the dataframe, that gives us a total sample size of 25 paired samples.

### Descriptive Statistics

With all of the observations saved into a dataframe, descriptive statistics can be computed from the dataset using R. Each treatment group, the congruent and incongruent trial groups, will be looked at seperately.

#### Congruent Trial

```{r echo=FALSE, Congruent_Summary}
summary(stroop$congruent)
print.default(sd(stroop$congruent))
```

The mean and median for the congruent trial data are 14.23 and 13.90 seconds. The median and mean are relatively close to each other, which may suggest that the data is evenly distributed and possibly normal. The standard devation is ~3.57.  

#### Incongruent Trial

```{r echo=FALSE, Incongruent_Summary}
summary(stroop$incongruent)
print.default(sd(stroop$incongruent))
```

Compared to the congruent trial data, the incongruent trial data has a much higher mean (21.86) and median (20.88). The standard deviation (~4.76) is also higher, meaning that there is more variation in the data compared to the congruent trial data. The incongruent trial data also has a very high maximum value.

### Visualizing the Dataset

Although descriptive statistics provide a good summary of the data, they do not provide insight into the data's shape or distribution. In order to learn more about the two trials, they need to be visualized.

Let's look at the congruent trial distribution first, visualized as a histogram:

```{r warning=FALSE, Congruent_Visual}
library(ggplot2)
library(gridExtra)

ggplot(data = stroop, 
       aes(x = congruent)) +
  geom_histogram(binwidth = 2, 
                 fill = "#538fef",
                 color = "black") +
  scale_x_continuous(breaks = c(seq(8, 36, 2))) +
  labs(x = "Time (s)", 
       y = "Frequency", 
       title = "Congruent Trial")
```

This distribution has a semi-normal shape with a majority of the values clustered near the median.

How does this plot compare to the incongruent trial?

```{r, Incongruent_Visual}
ggplot(data = stroop, 
       aes(x = incongruent)) +
  geom_histogram(binwidth = 2, 
                 fill = "#538fef",
                 color = "black") +
  scale_x_continuous(breaks = c(seq(8, 36, 2))) +
  labs(x = "Time (s)", 
       y = "Frequency", 
       title = "Incongruent Trial")
```

The incongruent trial data is right skewed, with a large gap between 26 and 34 seconds. These could potentially be outliers.

*****

Because of the different x-axis limits between the two plots, it is difficult to compare the two trials side by side. By plotting both visuals together using the same x limits, it should be much easier to compare them. The addition of a vertical line for the mean and a density curve should also improve comparison between the two trials.

```{r, Both_Visual}
c_vis <- ggplot(data = stroop, 
                aes(x = congruent)) +
           geom_histogram(binwidth = 2, 
                          fill = "#538fef",
                          color = "black") + 
           geom_density(aes(y = ..count.. * 2)) + ## scaled to better shape plot
           geom_vline(color = "#ed3838", 
                      xintercept = mean(congruent),
                      size = 1.25,
                      linetype = 2) +
           scale_x_continuous(limits = c(4, 40), 
                              breaks = c(seq(4, 40, 2))) +
           labs(x = "Time (s)", 
                y = "Frequency", 
                title = "Congruent Trial")

ic_vis <- ggplot(data=stroop, 
                 aes(x = incongruent)) +
            geom_histogram(binwidth = 2, 
                           fill = "#538fef",
                           color = "black") +
            geom_density(aes(y = ..count.. * 2)) + ##scaled here as well
            geom_vline(color = "#ed3838", 
                       xintercept = mean(incongruent),
                       size = 1.25,
                       linetype = 2) +
            scale_x_continuous(limits = c(4, 40), 
                               breaks = c(seq(4, 40, 2))) +
            labs(x = "Time (s)", 
                 y = "Frequency", 
                 title = "Incongruent Trial")

grid.arrange(c_vis, ic_vis)
```

This visualization really captures everything we need to compare the two distributions.

It is clear that the incongruent trial took longer to complete times on average than the congruent trial. But is this difference significant, or is the difference jsut due to chance? Only a statistical test can answer that.

### t-test and Conclusion

#### Pair Sample t-test Assumptions

In order to perform a paired sample t-test, there are [a couple requirements that must be met](https://libguides.library.kent.edu/SPSS/PairedSamplestTest):

1. Dependent variable must be continuous.
2. Samples/groups must be related (Ex: Person 1 takes both congruent and incongruent test)
3. Random sample from population
4. Approximately normal distribution of the difference between paired values
5. No outliers in difference between paired groups

To use the paired t-test on this dataset, we assume that the distribution of the difference between paired values  is normal. We also assume that the sample was taken randomly.

#### Calculating the t-score Manually

__The formula for the dependent t-test is as follows:__

$$
t = \frac{\overline{x}_{diff} - 0}{s_\overline{x}}
$$

__where:__

$$
s_\overline{x} = \frac{s_{diff}}{\sqrt{n}}
$$

__where:__

&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;$\overline{x}_{diff}$ = the difference between $\overline{x}_c$ and $\overline{x}_i$. \
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;$s_\overline{x}$ = the standard error of the mean. \
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;$s_{diff}$ = the standard deviation of the differences. \
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;$n$ = the sample size. \
  

Lets start by solving for $\overline{x}_{diff}$. In order to solve $\overline{x}_{diff}$, a seperate column for the difference between the data points in the congruent and incongruent trials is created.

```{r, Difference}
stroop$difference <- stroop$congruent - stroop$incongruent
```

The difference mean is then calculated and stored in `xdiff`:

```{r, xdiff}
xdiff <- mean(stroop$difference)
print(xdiff)
```

Now to find $s_\overline{x}$. To do that, $s_{diff}$ is calculated. `n` is also stored for later calculations:

```{r, SD_Diff}
sdiff <- sd(stroop$difference)
n <- 25

print(sdiff)
```

With 4.76 as $s_{diff}$ and 25 as $n$, $s_\overline{x}$ can now be calculated.

```{r, SE}
se <- sdiff / sqrt(n)
print(se)
```

All the pieces are in place. It is time to calculate the t-value:

```{r, t-test}
t_score <- xdiff / se
print(t_score)
```

Using the [t-table](https://s3.amazonaws.com/udacity-hosted-downloads/t-table.jpg), the __t-critical value__ is ±2.064, based upon a two-tailed test, 24 degrees of freedom, and a 0.05 confidence level.

The p-value can be calculated usin the built-in R function `pt`:

```{r}
format(2*pt(t_score, 24), digits = 1, scientific = FALSE)
```

#### Calculating t-score using R

In the previous section, _t_ was mathematically calculated by hand. This was done intentionally to show the math behind the t-score.

With R, a t-test can be run very easily using `t.test`:

```{r, r_ttest}
t.test(stroop$congruent, y=stroop$incongruent, paired =TRUE)
```

The results from the R t-test are identical to the results obtained manually.

#### Conclusion

Because the t-value (-8.36) is well below the t-critical value (-2.064), we reject the null hypothesis that the two groups are the same. The congruent and incongruent trials are significantly different at the 0.05 confidence level, with a p-value < .05.

The incongruent words took much longer to read on average than the congruent words. One theory is that the incongruent words slow the processing time of the brain as to tries to say the correct color. Because the incongruent words give conflicting input, it hampers the brains ability to choose the color quickly. The congruent words are encoded correctly with there corresponding color, allowing to brain to quickly identify and pronounce the color. 

### Reflection