---
title: 'Inferential Statistics: Exploration of the Stroop Effect'
author: "Tyler C Julian"
date: "February 19, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Independent and Dependent variables

The __independent variable__ for this experiment would be the the conditions of the words. They change from congruent to incongruent between the two trials.

The __dependent variable__ would the the amount of time it takes a participant to complete the test. This is measured in seconds. 

### Hypothesis and Statistical Test

The __null hypothesis__ for this experiment is that there will not be a statistical difference in the average time between the congruent and incongruent conditions. 

The __alternative hypothesis__ is that there will be a statistical difference in the average time between the congruent and incongruent conditions.

These hypotheses can be pictured below, with _H~0~_ being the null hypothesis, _H~a~_ being the alternative hypothesis, _£g~c~_, being the average congruent time, and _£g~i~_ being the average incongruent time:

\begin{align}
H_0 : £g_c = £g_i \\
H_a : £g_c ¡Ú £g_i
\end{align}

The statistical test that will be used for this experiment is a __two-tailed t-test__. This t-test is a dependent t-test with paired samples, and is chosen because the population parameters are unknown. It is two-tailed because we are unsure whether the times will rise or fall between word conditions.

### Descriptive Statistics

Descriptive statistics are pulled from [this dataset](https://drive.google.com/file/d/0B9Yf01UaIbUgQXpYb2NhZ29yX1U/view).

```{r}
congruent <- c(12.079, 16.791, 9.564, 8.63, 14.669, 12.238, 14.692, 8.987, 
               9.401, 14.48, 22.328, 15.298, 15.073, 16.929, 18.2, 12.13, 
               18.495, 10.639, 11.344, 12.369, 12.944, 14.233, 19.71, 16.004)

incongruent <- c(19.278, 18.741, 21.214, 15.687, 22.803, 20.878, 24.572, 17.394,
                 20.762, 26.282, 24.524, 18.644, 17.51, 20.33, 35.255, 22.158, 
                 25.139, 20.429, 17.425, 34.288, 23.894, 17.96, 22.058, 21.157)

stroop <- data.frame(congruent, incongruent)
```

```{r echo=FALSE}
summary(stroop$congruent)
print.default(sd(stroop$congruent))
```

```{r echo=FALSE}
summary(stroop$incongruent)
print.default(sd(stroop$incongruent))
```
### Visualizing the Dataset

### Two-Tailed T-Test and Conclusion

### Reflection